---
title: "STATS 500 HW1"
author: "Minxuan Chen"
date: last-modified
format:
  html:
    toc: true
    css: styles.css
    fig-align: "center"
    fig-width: 8
    fig-height: 6
    embed-resources: true
    format-links: false
    execute:
      warning: true
      freeze: auto
  pdf:
    documentclass: article
    papersize: a4paper
    toc: true
    number-sections: true
    colorlinks: true
    fig_width: 8
    fig_height: 6
    toc_depth: 2
    urlcolor: blue
    fontsize: 12pt
code-fold: show
code-overflow: scroll
code-line-numbers: true
---
Github repo: <https://github.com/PKUniiiiice/STATS_500>

```{r pre}
#| include: false
library(faraway)
library(ggplot2)
library(ggplot2)
library(cowplot)
library(dplyr)
library(GGally)
library(reshape)
library(dplyr)
```

## Problem 1
The dataset `teengamb` concerns a study of teenage gambling in Britain.       

<font size="+1">**Glimpse**</font>

```{r p1}
data(teengamb)
head(teengamb)
```

<font size="+1">**Descriptive Statistics**</font>

  Note that `sex` is a categorical variable so here we only provide descriptions for the other variables.
```{r p11}
summary(teengamb[,2:5])
```
  
It seems that the variable `gamble` is heavily right-skewed.

<font size="+1">**Distribution and Correlation**</font>

```{r p111}
md <- melt(teengamb[2:5], id.vars=NULL)
ggplot(md, aes(variable, value)) +
  geom_boxplot() +
  facet_wrap(~variable, scales="free")
ggpairs(teengamb[2:5])
```

In the boxplot, we can observe outliers in all variables except for `status`, with `gamble` having the highest number of outliers.

Regarding correlation, significant linear relationships are apparent between `status` and `verbal`, as well as between `income` and `gamble`. These relationships are consistent with the meanings of these variables.

<font size="+1">**Gender Differences**</font>

```{r p1111}
teengamb$sex <- as.factor(teengamb$sex)
md <- melt(teengamb, id.var=c("sex"), var="variables")
p <- ggplot(data=md, aes(x=variables, y=value, fill=sex))+
     geom_boxplot()+
     facet_wrap(~variables, scale="free")
show(p)
```

In the variables `status` and `gamble`, there appears to be noticeable differences between males and females. Males tend to have higher socioeconomic status scores as well as higher expenditures on gambling.

```{r p11111}
ggpairs(teengamb, columns=2:5, ggplot2::aes(color=sex))
```

Note that in the relationships `status`-`verbal` and `income`-`gamble`, there are noticeable differences between males and females (i.e. different corr coef). For instance, in the case of `income`-`gamble`, in males, teenagers with higher income tend to have higher expenditures on gambling. However, in females, gambling expenditure remains low, regardless of their income level. 

## Problem 2
### (a)
By definition, for $E(AZ)=AE(Z)$, we need to verify 
$$
[E(AZ)]_i = [AE(Z)]_i
$$
Note that
$$
\begin{aligned}[]
[E(AZ)]_i &= E([AZ]_i) \quad \text{(By def.)} \\
&=E\left(\sum_{j=1}^m A_{ij} Z_j \right) \\
&=\sum_{j=1}^m A_{ij} E(Z_j) \quad \text{(Linearity)} \\
&=\sum_{j=1}^m A_{ij} [E(Z)]_j \quad \text{(By def.)} \\
&=[AE(Z)]_i
\end{aligned}
$$

Similarly, for $\operatorname{Cov}(AZ)=A\operatorname{Cov}(Z) A^\top$, we need to verify 
$$
\operatorname{Cov}(AZ)_{ij} = [A \operatorname{Cov}(Z) A^{\top}]_{ij}
$$
Note that
$$
\begin{aligned}
LHS &= \operatorname{Cov}([AZ]_i, [AZ]_j) \\
&=\operatorname{Cov}\left( \sum_{s=1}^m A_{is} Z_s ,  \sum_{t=1}^m A_{jt} Z_t \right)\\
&=  \sum_{t=1}^m  \sum_{s=1}^m   A_{is}   A_{jt}  \operatorname{Cov}(Z_s,Z_t) \\
&= \sum_{t=1}^m  \sum_{s=1}^m   A_{is}   A_{jt}  \operatorname{Cov}(Z)_{st} 
\end{aligned}
$$
and
$$
\begin{aligned}
RHS &= \left[A \operatorname{Cov} (Z) A^\top\right]_{ij} \\
    &= \sum_{t=1}^m  [A \operatorname{Cov} (Z)]_{it} \, [A^\top]_{tj} \\
    &= \sum_{t=1}^m \left(  \sum_{s=1}^m A_{is} \operatorname{Cov} (Z)_{st}    \right) [A^\top]_{tj}   \\
    &=  \sum_{t=1}^m \left(  \sum_{s=1}^m A_{is} \operatorname{Cov} (Z)_{st}    \right) A_{jt} \quad (  A_{jt} = [A^\top]_{tj}) \\
    &= \sum_{t=1}^m  \sum_{s=1}^m   A_{is}   A_{jt}  \operatorname{Cov}(Z)_{st} \\
    &=LHS
\end{aligned}
$$

### (b)
We have
$$
Y=\left[\begin{array}{c}
Z_1+2 Z_2 \\
Z_1-2 Z_2 \\
-Z_1+Z_2
\end{array}\right] = \left[ \begin{array}{rr} 1&2 \\ 1&-2 \\ -1 & 2
\end{array} \right] 
\left[\begin{array}{c}
Z_1 \\
Z_2 
\end{array}\right] = AZ
$$
 and
 $$
 \operatorname{Cov} (Z) = \left[ \begin{array}{rr} 3&-1 \\ -1&3 
\end{array} \right] 
 $$
Therefore,
$$
\operatorname{Cov} (Y) = A  \operatorname{Cov} (Z) A^\top
$$
```{r p2b}
A <- matrix(c(1,1,-1,2,-2,2),ncol=2)
cov.z <- matrix(c(3,-1,-1,3),ncol=2)
cov.Y <- A%*%cov.z%*%t(A)
cov.Y
```
and
$$
\operatorname{Corr} (Y) = [\text{diag}( \operatorname{Cov} (Y))]^{-\frac{1}{2}}   \operatorname{Cov} (Y) [\text{diag}( \operatorname{Cov} (Y))]^{-\frac{1}{2}}
$$

```{r p2bb}
diag(diag(cov.Y)^(-1/2))%*%cov.Y%*%diag(diag(cov.Y)^(-1/2))
```

## Problem 3
It's known that
$$
s_y = \sqrt{ \frac{1}{n} \sum_{i=1}^n\left(y_i-\bar{y}\right)^2}, s_x = \sqrt{ \frac{1}{n} \sum_{i=1}^n\left(x_i-\bar{x}\right)^2}
$$
$$
r = \frac{\frac{1}{n} \sum_{i=1}^n (x_i-\bar{x})(y_i-\bar{y})  }{s_x \cdot s_y}
$$
We go from $y=\hat{\beta_0}+\hat{\beta_1}x$
$$
\begin{aligned}
y &= \bar{y}-\hat{\beta_1} \bar{x} + \hat{\beta}_1x\\
  &= \bar{y} + \hat{\beta}_1 (x-\bar{x}) \\
  &=   \bar{y} + \frac{\sum_{i=1}^n\left(x_i-\bar{x}\right)\left(y_i-\bar{y}\right)}{\sum_{i=1}^n\left(x_i-\bar{x}\right)^2} (x-\bar{x}) \\ 
  &= \bar{y} + \frac{ n\cdot r \cdot s_x \cdot s_y } {\sum_{i=1}^n\left(x_i-\bar{x}\right)^2} (x-\bar{x}) \\
  &= \bar{y} + \frac{ r \cdot s_x \cdot s_y } {\frac{1}{n}\sum_{i=1}^n\left(x_i-\bar{x}\right)^2} (x-\bar{x}) \\
  &= \bar{y} + \frac{ r \cdot s_x \cdot s_y }{s_x^2}(x-\bar{x}) \\
\end{aligned}
$$
Therefore, we have
$$
\frac{y-\bar{y}}{s_y} = r\cdot \frac{x-\bar{x}}{s_x}
$$

