---
title: "STATS 500 HW10"
author: "Minxuan Chen"
date: last-modified
format:
  html:
    toc: true
    css: styles.css
    fig-align: "center"
    fig-width: 6
    fig-height: 5
    embed-resources: true
    format-links: false
    execute:
      warning: true
      freeze: auto
  pdf:
    papersize: a4paper
    pdf-engine: pdflatex
    toc: false
    number-sections: false
    colorlinks: true
    fig-width: 6
    fig-asp: 0.8
    urlcolor: blue
    fontsize: 12pt
    geometry:
      - top=30mm
      - left=15mm
      - right=15mm
      - bottom=30mm
      - heightrounded
code-fold: show
code-overflow: scroll
code-line-numbers: true
---

```{=latex}
% Trigger ToC creation in LaTeX
\tableofcontents
\thispagestyle{empty}
\vspace{20pt}
Github repo: \url{https://github.com/PKUniiiiice/STATS_500}

\newpage

\setcounter{page}{1}
```

## Problem 1
### (a)
```{r}
library(faraway)
data(seeds)

par(mfrow=c(1,2))
plot(seeds[seeds$covered=='no', 2],
     seeds[seeds$covered=='no', 1],
     ylab='Germination percentage', xlab='Moisture level',
     main='Uncovered boxes', xaxt='n'
     )
axis(1,at=seq(1,11,by=2))
plot(seeds[seeds$covered!='no', 2],
     seeds[seeds$covered!='no', 1],
     ylab='Germination percentage', xlab='Moisture level',
     main='Covered boxes',xaxt='n')
axis(1,at=seq(1,11,by=2))

```

We find that, in both uncovered and covered boxes, the germination percentage seems to increase with moisture level when the level is relative low (<5), and the germination percentage seems to decrease with moisture level when the level is relative high (>5). And if the moisture level is in a relative high level (=7, =9). If the level is too high (=11), no seed can germinate.

So, considering the whole range of moisture level, the relationship seems in a quadratic form with a negative leading term.

### (b)
```{r}
seeds$box <- factor(rep(1:8,each=6))

library(ggplot2)
# Create the ggplot objects
plot_uncovered <- ggplot(subset(seeds, covered == 'no'), aes(x = moisture, y = germ, color = box, group=box)) +
  geom_line() +
  geom_point() +
  labs(title = 'Uncovered boxes', x = 'Moisture level', y = 'Germination percentage') +
  scale_x_continuous(breaks = seq(1, 11, 2))

plot_covered <- ggplot(subset(seeds, covered != 'no'), aes(x = moisture, y = germ, color = box, group=box)) +
  geom_line() +
  geom_point() +
  labs(title = 'Covered boxes', x = 'Moisture level', y = 'Germination percentage') +
  scale_x_continuous(breaks = seq(1, 11, 2))

# Arrange the plots side by side
gridExtra::grid.arrange(plot_uncovered, plot_covered, ncol = 2)
```

There is no indication of a box effect. Since we observe the same tendency among different boxes (in uncovered boxes or in covered boxes.)

### (c)
In (a), we observe a quadratic relationship between moisture and germination percentage.

Therefore, we can add a quadratic term of moisture. That is
```{r}
#dropna
seeds <- na.omit(seeds)

logitm <- glm(cbind(germ, 100-germ) ~ moisture + I(moisture^2) + box + covered,
              family=binomial(link=logit), data=seeds)
summary(logitm)
```
<!--We can also fit a probit model to compare.

```{r}
probitm <- glm(cbind(germ, 100-germ) ~ moisture + I(moisture^2) + box,
              family=binomial(link=probit), data=seeds)
summary(probitm)
```
-->

We find that the p-value of moisture squared is strongly significant, so it's reasonable to add this term.

Moreover, the effect of `coverage` is not identifiable since the box number and coverage are completely correlated, i.e. we can tell coverage or not totally from box number (1,2,3,4=uncovered, 5,6,7,8=covered).

To solve this problem, we may need to remove the `box` predictor from the model. 

We draw a plot of the predictions.
```{r}
xnew <- data.frame(
  moisture = rep(seq(1,11,0.5),8),
  covered = factor(rep(c('no','yes'),each=84)),
  box = factor(rep(1:8, each=21))
)
pred.logit <- predict(logitm, xnew, type='response')
temp <- cbind(xnew, pred.logit)

plot_uncovered <- ggplot(subset(temp, covered == 'no'), aes(x = moisture, color=box)) +
 # geom_point(aes(x=seeds$moisture, y = seeds$germ))+
  geom_line(aes(y = pred.logit*100,), linewidth = 0.5,
            linetype=2) +
  geom_point(data=subset(seeds, covered == 'no'),
             aes(x=moisture, y = germ))+
  labs(title = 'Uncovered boxes', x = 'Moisture level', y = 'Germination percentage') +
  scale_x_continuous(breaks = seq(1, 11, 2))

plot_covered <- ggplot(subset(temp, covered != 'no'), aes(x = moisture, color=box)) +
 # geom_point(aes(x=seeds$moisture, y = seeds$germ))+
  geom_line(aes(y = pred.logit*100,), linewidth = 0.5,
            linetype=2) +
  geom_point(data=subset(seeds, covered != 'no'),
             aes(x=moisture, y = germ))+
  labs(title = 'Covered boxes', x = 'Moisture level', y = 'Germination percentage') +
  scale_x_continuous(breaks = seq(1, 11, 2))

# Arrange the plots side by side
gridExtra::grid.arrange(plot_uncovered, plot_covered, ncol = 2)

```

### (d)
To test the significance of a box effect in the model, we remove the box predictor and use anova.
```{r}
logitm2 <- glm(cbind(germ, 100-germ) ~ moisture + I(moisture^2) + covered,
              family=binomial(link=logit), data=seeds)
anova(logitm2, logitm)
```
The deviance is 3.2148, by $D \approx \chi_{n-s}^2$,
the p-value is
```{r}
pchisq(3.2148, df=6, lower.tail=F)
```
The p-value is larger than 0.05, so we conclude that we do not reject the null model. That is, we accept the model without box effect.

If using Pearson’s Chi-squared test, we need Pearson residuals.
```{r}
resM2 <- residuals(logitm2, type='pearson')
resM1 <- residuals(logitm, type='pearson')
pchisq(sum(resM2^2)-sum(resM1^2), df=6, lower.tail=F)
```
The p-value is still larger than 0.05, so we accept the model without box effect.

### (e)
```{r}
#get prediction, full model is logitm
seeds$pred <- predict(logitm, type='response')
```
In uncovered boxes
```{r}
seeds.no <- subset(seeds, covered=='no')
seeds.no[seeds.no$pred==max(seeds.no$pred),]
```
The predicted maximum germination for
uncovered boxes occurs at moisture level=5.

In covered boxes
```{r}
seeds.yes <- subset(seeds, covered=='yes')
seeds.yes[seeds.yes$pred==max(seeds.yes$pred),]
```
The predicted maximum germination for
uncovered boxes also occurs at moisture level=5.

### (f)
```{r}
plot(logitm, 1)
```

It seems the distribution of residuals is uneven across the predicted values. In other words, there may be non-constant variance in errors. And there is also slightly non-linear in the residuals since the red line is not horizontal.

### (g)
```{r}
seeds$resid <- resid(logitm)

ggplot(seeds, aes(x = moisture, y=resid, color=covered)) +
  geom_point()+
  geom_smooth(formula=y~x, method="loess", se=FALSE, linetype = 2, lindwidth=0.5)+
  labs(title = 'Residuals vs Moisture',
       x = 'Moisture level',
       y = 'Residuals') +
  scale_x_continuous(breaks = seq(1, 11, 2))
```

From this plot, we can still observe non-linear pattern in the residuals. Therefore, the moisture squared tern still does not fit the data well. We may add some higher terms as a trial.

## Problem 2
We estimate the dispersion parameter by
$$
\hat{\sigma}^2 = \frac{X^2}{n-p}
$$
that is
```{r}
sigma2 <- sum(resid(logitm, type='pearson')^2)/logitm$df.residual
sigma2
sumary(logitm,dispersion=sigma2)
drop1(logitm, scale=sigma2, test="F")
```
The estimated dispersion parameter value is 13.77, which is much larger than the assumption of 1 we used in last problem.

From the summary and F-test results, we find the standard error and p-value of coefficients changed. But the conclusion of no box effect still holds.

## Problem 3
First we consider the binomial model.

The deviation is defined as
$$
D=2 \sum_{i=1}^n\left[y_i \log \frac{y_i}{\hat{y}_i}+\left(n_i-y_i\right) \log \frac{n_i-y_i}{n_i-\hat{y}_i}\right], \hat{y}_i = n_i\hat{p}_i
$$

We can rewrite the deviance as
$$
\begin{aligned}
D&=2 \sum_{i=1}^n\left[y_i \log \frac{y_i}{\hat{y}_i}+\left(n_i-y_i\right) \log \frac{n_i-y_i}{n_i-\hat{y}_i}\right]\\
&=2 \sum_{i=1}^n\left[y_i \log y_i - y_i \log n_i\hat{p}_i +\left(n_i-y_i\right) \log \left(n_i-y_i\right)  - \left(n_i-y_i\right) \log \left(n_i-n_i\hat{p}_i\right) \right]\\
&=-2 \sum_{i=1}^n y_i (\log n_i+\log\hat{p}_i)+ \left(n_i-y_i\right) (\log n_i +\log  \left(1-\hat{p}_i\right)) + C_1(n_i,y_i)\\
&=-2  \sum_{i=1}^n y_i \log\hat{p}_i+ \left(n_i-y_i\right)\log  \left(1-\hat{p}_i\right)  +  C_2(n_i,y_i) \\
\end{aligned}
$$
in which $C_1(n_i,y_i), C_2(n_i,y_i)$ are terms that only depend on $n_i, y_i, i=1,2,...n$.

AIC of binomial model is defined as
$$
\begin{aligned}
AIC &= -2\log (\hat{L}) +2q \\
&=-2 \sum_{i=1}^n\left[\log \left(\begin{array}{l}
n_i \\
y_i
\end{array}\right)+y_i \log \hat{p}_i+\left(n_i-y_i\right) \log \left(1-\hat{p}_i\right)\right] + 2q \\
&= -2 \sum_{i=1}^n y_i \log \hat{p}_i+\left(n_i-y_i\right) \log \left(1-\hat{p}_i\right)  + 2q + C_3(n_i,y_i)
\end{aligned}
$$

Comparing these two expressions, we find that
$$
AIC^* = D+2q = AIC - C_3(n_i,y_i) + C_2(n_i,y_i)
$$
Note that the difference between AIC and AIC* is a term that only depend on $n_i, y_i, i=1,2,...n$, which is the same across all models. Therefore, minimize AIC* criteria is equivalent to minimize AIC criteria.

Next, for binary model, deviation is defined as
$$
D = -2 \sum_{i=1}^n\left[y_i \log \left(\hat{p}_i\right)+\left(1-y_i\right) \log \left(1-\hat{p}_i\right)\right]
$$

Note that in binomial model, if we take all $n_i=1$, then the deviance will become the same as that in binary situation except for a constant difference, which is independent of model and only depends on data.
$$
\begin{aligned}
&D= -2  \sum_{i=1}^n y_i \log\hat{p}_i+ \left(1-y_i\right)\log  \left(1-\hat{p}_i\right)+C_2(n_i=1,y_i)
\end{aligned}
$$
Moreover, for AIC of binary model, we can take all $n_i=1$ in AIC of binomial model to get the result.

Therefore, for both deviance and AIC, binary model is a special case of binomial model. So previous conclusion still holds.

From above, selecting the model via AIC criteria can be done by picking the model
that minimizes the AIC* criteria

